Tree establishment in models of forest dynamics: a key priority for further research

### 2023-03-11 manuscript/manuscriptRW.Rmd
re-define questions of the paper in the introduction

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
nocite: |
  @reyer, @bugmann1996a, @bohn2014, @fabrika2005, @mette2014, @smith2001a, @seidl2012a, @lischke2006, @schumacher2006, @scheiter2009, @scheller2004, @morin2021
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

More specifically, we aimed to answer the following questions. 1) Are
models of forest dynamics capturing accurately tree ingrowth levels
across environmental gradients? 2) Are they matching the initial species
composition of the tree ingrowth layer? And 3), do they correctly
represent the mortality in tree establishment?

NEW

@@

2)  Species specific recruitment values in ennvironmental gradients?

3)  Are the models traits linked to the tree establishment values?

environmental gradients?

2)  Are they matching the initial species composition of the tree
    ingrowth layer? And

3), do they correctly represent the mortality in tree establishment?

@@

modelDF <- readr::read_csv(here::here("data", "modelsDescription.csv"))
modelDF$`Species simulated`[modelDF$`Species simulated` == "F. sylbatica, P. sylvestris, P. abies, A. alba"] <- "F. sylbatica,  P. sylvestris, P. abies,  A. alba"
modelDF$Sampling <- NULL

#there is a better way of doing footnotes, have a look here https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf and change
names(modelDF)[2] <- paste0(names(modelDF)[2], "(a)")
names(modelDF)[3] <- paste0(names(modelDF)[3], "(b)")
names(modelDF)[4] <- paste0(names(modelDF)[4], "(c)")
names(modelDF)[5] <- paste0(names(modelDF)[5], "(d)")
names(modelDF)[6] <- paste0(names(modelDF)[6],"(e)")
names(modelDF)[11] <- paste0(names(modelDF)[11],"(f)")

kable(modelDF, 
      #format = "markdown",
      format = "latex", booktabs = TRUE,
      caption = "List of models included.") |>
        kableExtra::landscape() |>
        kableExtra::kable_styling(full_width = F, latex_options = "scale_down") |>
    # fix width by column
      kableExtra::column_spec(column = 1, width = "5em") |> 
      kableExtra::column_spec(column = 2, width = "3em") |> 
      kableExtra::column_spec(column = 3, width = "3em") |> 
      kableExtra::column_spec(column = 4, width = "4em") |>
      kableExtra::column_spec(column = 5, width = "5em") |>
      kableExtra::column_spec(column = 6, width = "5em") |>
      kableExtra::column_spec(column = 7, width = "5em") |>
      kableExtra::column_spec(column = 8, width = "8em") |>
      kableExtra::column_spec(column = 9, width = "8em") |>
      kableExtra::column_spec(column = 10, width = "5em") |>
      kableExtra::column_spec(column = 11, width = "5em") |>
      kableExtra::column_spec(column = 12, width = "8em") |>
      kableExtra::add_footnote(c("S-Stand, L-Landscape, G-Global",
                                 "PB-Process based, EM-Empirical", 
                                 "C-Cohort, I-Individual",
                                 "RG-Regeneration, RC-Recruitment",
                                 "Y-Yes, N-No",
                                 "D-Daily, M-Monthly"),

@@

```{r map, echo=FALSE, fig.cap="The 200 sites of the EuFoRIa network that were used in the present study, showing the dominant tree species per site in the observed ingrowth at the 10 cm threshold.", out.width='80%'}
knitr::include_graphics(here::here("figures", "theMap.png"))
```

@@

```{=tex}
\begin{equation}
H_{n,r} = - \sum_{i = 1}^{s}{p_{n,i}ln(p_{n,r,i})}\label{eqpiShannonR}
\end{equation}
```
```{=tex}
\begin{equation}
p_{n,r,i} = \frac{r_{BA_{n,i}}}{TotalR_{BA_{n}}} \label{eqpiShannonR2}
\end{equation}
```
```{=tex}
\begin{equation}
H_{n,s} = - \sum_{i = 1}^{s}{p_{n,i}ln(p_{n,i})}\label{eqpiShannon}
\end{equation}
```
```{=tex}
\begin{equation}
p_{n,s,i} = \frac{BA_{n,i}}{Total_{BA_{n}}} \label{eqpiShannon2}
\end{equation}
```
where $s$ is the total number of species present and have basal area
above zero in sample $n$; $p_{n,r,i}$ is the proportion of the species
$i$ calculated as the ingrowth basal area ($r_{BA_{n,i}}$) for that
species $i$ relative to the recruited basal area ($TotalR_{BA_{n}}$) of
the sample $n$; $p_{n,s,i}$ is the proportion of the species $i$
calculated as the basal area ($BA_{n,i}$) for that species $i$ relative
to the total basal area ($TotalR_{BA_{n}}$) of the sample $n$.

@@

```{r over, echo=FALSE, fig.cap="Mean ingrowth rates across all samples per site, plotted for the 200 sites and for each model. The red dotted lines shows the 2.5 and 97.5 percentiles for the 7 cm diameter threshold in the observed data. There are two boxplots for each model where the left lighted boxplot corresponds to 7cm and the right darker boxplot to 10cm", out.width='100%'}
        knitr::include_graphics(here::here("figures", "recOverAll710.png"))
```

@@

The models did not match the observed ingrowth patterns across gradients
of soil moisture and temperature (Figures S1 and S2), and neither was
there a clear pattern across models. Regarding the soil moisture
gradient (Figure S1), a group of models featured decreasing ingrowth
with increasing soil moisture (4C, ForClim 1, ForClim11, SIBYLA, xComp,
PICUS, LandClim, TreeMig and iLand), while a few models showed the
opposite trend (ForCEEPS(f), FORMIND and LPJ-GUESS). Across the
temperature gradient (Figure S2), all the models that featured a
decrease in ingrowth rates with increasing soil moisture showed the
opposite trend (i.e., more ingrowth with increasing temperature), except
for 4C. The other models did not feature a clear pattern.

@@

```{r H710, echo=FALSE, fig.cap="Shannon index of tree ingrowth, calculated by basal area, as the mean value across the 200 samples per site for the diameter thresholds of 7 and 10 cm, respectively. The red dotted lines represent the 20th percentile of the observed data for the threshold of 7 cm and the 70th percentile of the observed data for the threshold of 10 cm.", out.width='100%'}
    knitr::include_graphics(here::here("figures", "H7_10.png"))
```

@@

Lastly, there is no model that performed well across the five species
regarding the species-specific ingrowth rates in the environmental space
of Figure 6. Some models (e.g., ForClim11, PICUS) tended to consistently
overestimate the ingrowth share of all five species, but most models
overestimated the ingrowth share of some species while underestimating
the share of others (e.g., ForCEEPS, LPJ-GUESS).

```{r diff, echo=FALSE, fig.cap="Ingrowth rates of the five main species in terms of their basal area share in the observations (top row) and the performance of each model across the environmental gradients (other rows) for the 7 cm ingrowth threshold. The values shown are the mean of the 200 samples per site and across the sites in each bin (tile), with ten bins per gradient. The sizes of the circles represent the ratio between the ingrowth basal area of the species and the total ingrowth basal area for all species. The absence of a circle indicates a zero basal area share in the ingrowth, or the absence of ingrowth altogether. The color gradient (for the models only) shows the difference between the simulated and observed ratio of ingrowth basal area of the species and the total ingrowth basal area for all the species.", out.width='100%', out.height= '85%'}
knitr::include_graphics(here::here("figures", "HeatEnvGrad_10circles_7.png"))
```

@@

There were strong differences among the models regarding the mortality
rate between 7 and 10 cm (Figure \@ref(fig:mortality11)). Most of the
models featured a mortality rate significantly larger than the observed
data, thus at least partially compensating for the general
overestimation of ingrowth rates (Figure \@ref(fig:over) and Table S5);
six models did not feature a significant difference compared to the
observed data (ForCEEPS, FORMIND, ForClim 1 and 11, and aDGVM2), and
only the model LPJ-GUESS had a significantly lower mortality rate. Four
models (xComp, PICUS, LandClim and TreeMig) featured very high mortality
rates (i.e., well above the Reineke self-thinning line), which
compensated for the strong initial overestimation of ingrowth (cf.
Figure 2). Interestingly, two models that underestimated overall
ingrowth levels (SIBYLA and Landis-II; cf. Figure \@ref(fig:over))
featured mortality rates that were close to but still above the
self-thinning line (Figure \@ref(fig:mortality11)).

@@

```{r mortality13, echo=FALSE, fig.cap="Ratio of tree ingrowth between the 7 and 10 cm thresholds (ingrowth 7 cm / ingrowth 10 cm) and the overestimation proportion at 7 cm ([simulated -observed] / observed) for the mean ingrowth per model across sites and samples. The horizontal dotted lines show a ratio equal to 1, indicating no decrease of ingrowth between 7 and 10 cm, and a ratio equal to 1.77 corresponding to the Reineke self-thinning line under even-aged conditions. The vertical line indicates zero overestimation.", out.width='80%'}
knitr::include_graphics(here::here("figures", "mort7_10_medianOver.png"))
```

>>>>

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
nocite: |
  @bugmann1996a, @bohn2014, @fabrika2005, @mette2014, @smith2001a, @langan2017, @seidl2012a, @lischke2006, @schumacher2006, @scheiter2009, @scheller2004, @morin2021, @lasch-born2020, @hickler2012, @fischer_lessons_2016
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

More specifically, we aimed to answer the following questions:

@@

2)  Do model traits explain differences in model performance?

3)  How well do models capture total ingrowth and individual species
    regeneration niches across environmental gradients of light
    availability, temperature, and soil moisture?

@@

modelDF <- readr::read_csv(here::here("data", "modelsDescription.csv"))
modelDF$Sampling <- NULL

#footnotes
names(modelDF)[2] <- paste0(names(modelDF)[2], "(a)")
names(modelDF)[3] <- paste0(names(modelDF)[3], "(b)")
names(modelDF)[4] <- paste0(names(modelDF)[4], "(c)")
names(modelDF)[5] <- paste0(names(modelDF)[5], "(d)")
names(modelDF)[6] <- paste0(names(modelDF)[6], "(e)")
names(modelDF)[11] <- paste0(names(modelDF)[11],"(f)")

kable(modelDF, 
      format = "latex", booktabs = TRUE,
      caption = "List of models included.") |>
        kableExtra::landscape() |>
        kableExtra::kable_styling(full_width = F, latex_options = "scale_down") |>
    # fix width by column
      kableExtra::column_spec(column = 1, width = "5em") |> 
      kableExtra::column_spec(column = 2, width = "3em") |> 
      kableExtra::column_spec(column = 3, width = "3em") |> 
      kableExtra::column_spec(column = 4, width = "4em") |>
      kableExtra::column_spec(column = 5, width = "5em") |>
      kableExtra::column_spec(column = 6, width = "5em") |>
      kableExtra::column_spec(column = 7, width = "5em") |>
      kableExtra::column_spec(column = 8, width = "8em") |>
      kableExtra::column_spec(column = 9, width = "8em") |>
      kableExtra::column_spec(column = 10, width = "5em") |>
      kableExtra::column_spec(column = 11, width = "5em") |>
      kableExtra::column_spec(column = 12, width = "8em") |>
      kableExtra::add_footnote(c("S-Stand, L-Landscape, G-Global",
                                 "PB-Process based, EM-Empirical", 
                                 "C-Cohort, I-Individual",
                                 "RG-Regeneration, RC-Recruitment",
                                 "Y-Yes, N-No",
                                 "D-Daily, M-Monthly"),

@@

```{r map, echo=FALSE, fig.cap="The 200 sites of the EuFoRIa network used in the present study, the colors show the dominant tree species per site in the observed ingrowth at the 10 cm threshold.", out.width='80%'}
knitr::include_graphics(here::here("figures", "theMap.png"))
```

@@

```{=tex}
\begin{equation}
H_{n,r} = - \sum_{i = 1}^{s}{p_{n,i}\ln(p_{n,r,i})}\label{eqpiShannonR}
\end{equation}
```
```{=tex}
\begin{equation}
p_{n,r,i} = \frac{r_{BA_{n,i}}}{TotalR_{BA_{n}}} \label{eqpiShannonR2}
\end{equation}
```
```{=tex}
\begin{equation}
H_{n,s} = - \sum_{i = 1}^{s}{p_{n,i}\ln(p_{n,i})}\label{eqpiShannon}
\end{equation}
```
```{=tex}
\begin{equation}
p_{n,s,i} = \frac{BA_{n,i}}{Total_{BA_{n}}} \label{eqpiShannon2}
\end{equation}
```
where $s$ is the total number of species present and have basal area
above zero in sample $n$; $p_{n,r,i}$ is the proportion of the species
$i$ calculated as the ingrowth basal area ($r_{BA_{n,i}}$) for that
species $i$ relative to the recruited basal area ($TotalR_{BA_{n}}$) of
the sample $n$; $p_{n,s,i}$ is the proportion of the species $i$
calculated as the basal area ($BA_{n,i}$) for that species $i$ relative
to the total basal area ($TotalR_{BA_{n}}$) of the sample $n$.

@@

```{r over, echo=FALSE, fig.cap="Mean ingrowth levels across all samples per site, plotted for the 200 sites and for each model. The red dashed lines shows the 25th and 75th percentiles for the 7 cm diameter threshold in the observed data. There are two boxplots for each model where the left lighted boxplot corresponds to 7cm and the right darker boxplot to 10cm", out.width='100%'}
        knitr::include_graphics(here::here("figures", "recOverAll710.png"))
```

@@

The models did not match the observed ingrowth patterns across gradients
of soil moisture and temperature (Figures S1 and S2), and neither was
there a clear pattern across models. Regarding the soil moisture
gradient (Figure S1), a group of models featured decreasing ingrowth
with increasing soil moisture (4C, ForClim 1, ForClim11, SIBYLA, xComp,
PICUS, LandClim, TreeMig and iLand), while a few models showed the
opposite trend (ForCEEPS(f), FORMIND and LPJ-GUESS). Across the
temperature gradient (Figure S2), all the models that featured a
decrease in ingrowth levels with increasing soil moisture showed the
opposite trend (i.e., more ingrowth with increasing temperature), except
for 4C. The other models did not feature a clear pattern.

@@

```{r H710, echo=FALSE, fig.cap="Shannon index of tree ingrowth, calculated by basal area, as the mean value across the 200 samples per site for the diameter thresholds of 7 and 10 cm, respectively. The red dashed lines represent the 25th percentile of the observed data for the threshold of 7 cm and the 75th percentile of the observed data for the threshold of 10 cm.", out.width='100%'}
    knitr::include_graphics(here::here("figures", "H7_10.png"))
```

@@

Lastly, there is no model that performed well across the five species
regarding the species-specific ingrowth levels in the environmental
space of Figure 6. Some models (e.g., ForClim11, PICUS) tended to
consistently overestimate the ingrowth share of all five species, but
most models overestimated the ingrowth share of some species while
underestimating the share of others (e.g., ForCEEPS, LPJ-GUESS).

```{r diff, echo=FALSE, fig.cap="Ingrowth levels of the five main species in terms of their basal area share in the observations (top row) and the performance of each model across the environmental gradients (other rows) for the 7 cm ingrowth threshold. The values shown are the mean of the 200 samples per site and across the sites in each bin (tile), with ten bins per gradient. The sizes of the circles represent the ratio between the ingrowth basal area of the species and the total ingrowth basal area for all species. The absence of a circle indicates a zero basal area share in the ingrowth, or the absence of ingrowth altogether. The color gradient (for the models only) shows the difference between the simulated and observed ratio of ingrowth basal area of the species and the total ingrowth basal area for all the species.", out.width='100%', out.height= '85%'}
knitr::include_graphics(here::here("figures", "HeatEnvGrad_10circles_7.png"))
```

@@

There were strong differences among the models regarding the mortality
rate between 7 and 10 cm (Figure \@ref(fig:mortality11)). Most of the
models featured a mortality rate significantly larger than the observed
data, thus at least partially compensating for the general
overestimation of ingrowth levels (Figure \@ref(fig:over) and Table S5);
six models did not feature a significant difference compared to the
observed data (ForCEEPS, FORMIND, ForClim 1 and 11, and aDGVM2), and
only the model LPJ-GUESS had a significantly lower mortality rate. Four
models (xComp, PICUS, LandClim and TreeMig) featured very high mortality
rates (i.e., well above the Reineke self-thinning line), which
compensated for the strong initial overestimation of ingrowth (cf.
Figure 2). Interestingly, two models that underestimated overall
ingrowth levels (SIBYLA and Landis-II; cf. Figure \@ref(fig:over))
featured mortality rates that were close to but still above the
self-thinning line (Figure \@ref(fig:mortality11)).

@@

```{r mortality13, echo=FALSE, fig.cap="Ratio of tree ingrowth between the 7 and 10 cm thresholds (ingrowth 7 cm / ingrowth 10 cm) and the overestimation proportion at 7 cm ([simulated -observed] / observed) for the mean ingrowth per model across sites and samples. The horizontal dashed lines show a ratio equal to 1, indicating no decrease of ingrowth between 7 and 10 cm, and a ratio equal to 1.77 corresponding to the Reineke self-thinning line under even-aged conditions. The vertical line indicates zero overestimation.", out.width='80%'}
knitr::include_graphics(here::here("figures", "mort7_10_medianOver.png"))
```
